---
title: "Geospatial Analysis of a Changing Road Network, 11-17-2025"
author: "Will Jones"
editor: source
format:
  html:
    embed-resources: true
execute: 
  echo: false
  cache: true
  warning: false
  message: false
  publish:
  resources:
  - "*.png"
  site-url: "https://willj98.quarto.pub/roads-project-07-28"
bibliography: references.bib
---

# Methods

```{r}
#| warning: false
#| echo: false
#| message: false
library(sf)
library(terra)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(nngeo)
library(leaflet)
library(leaflegend)
library(tigris)
library(foreign)
library(purrr)
library(tidyverse)
library(readr)
library(ggridges)
library(tmap)
library(landscapemetrics)
library(patchwork)
library(ggthemes)
library(gt)
```

### Data Acquisition and Preprocessing

Six primary spatial datasets were used in our analysis. The 2023 USGS National Land Cover Database (NLCD), containing 11 distinct land cover classes, was used to extract surface type at each of our sampled points [@usgs]. Likewise, attribute data for each sampled point was extracted using the EPA's ecoregion level IV map, watershed membership

US Census Bureau's TIGER/Line vector files were used as the road networks in this project. The most recently available 2024 road network data for the conterminous United States was downloaded in bulk using the `roads()` function from the Python package `pygris` [@pygris].

Comparable data for the year 2000 was downloaded from the archived Census FTP site `(https://www2.census.gov/geo/tiger/tiger2k/)`, and a multi-step data preprocessing workflow was performed before the data was implemented in the analysis workflow. First, `beautifulsoup4` was used to download vector dataset in `RT1` format, the historical file format for census geographic data [@beautifulsoup]. Using GDAL command-line utilities, files were converted to modern shapefile format, then merged into a single dataset using `geopandas` in Python [@geopandas]. Finally, feature correction was performed using geometry-snapping and densification processing using QGIS 3.36 "Maidenhead." To rectify data quality issues, namely feature accuracy and spatial mismatch between existing roads in the 2000 and 2024 datasets, the historical vector dataset was densified at regular 25m intervals, then snapped to the nearest matching feature in the current roads dataset. This was done to avoid erroneous measurement differences in distance to the nearest road for each year that may be incorrectly attributed to removal or addition of a new road feature.

`EPSG:5070, NAD 83 CONUS Albers | Unit: m`

```{r}
#| echo: false
#| message: false
#| include: false

# This table is easy to use for calculating summary statistics of each distribution
setwd("E:\\01_PROJECTS\\01_Roads\\06_fixing_final_files")
all_data <- st_read('pairwise_data_07_28.gpkg')
all_data <- all_data %>% mutate(ratio = (dist_24 / dist_00)) %>% mutate(log_ratio = log10(ratio))
#head(all_data)
# While we pivot this data to "long" format for easier visualization
long_data <- all_data %>%
  pivot_longer(
    cols = c(dist_00, dist_24, log_ratio),
    names_to = "measure_type", 
    values_to = "meters" 
  )
```

```{r}
#| echo: false
#| 
summary_table <- long_data %>% st_drop_geometry() %>% filter(measure_type != 'log_ratio') %>% group_by(measure_type) %>% summarise(Mean = mean(meters), 
          SD = sd(meters),
          Min = min(meters, na.rm = TRUE),
          `1st quartile` = quantile(meters, 0.25),
          Median = quantile(meters, 0.5),
          `3rd Quartile` = quantile(meters, 0.75),
          Max = max(meters),)

summary_table <- summary_table %>% 
  rename(Year = measure_type) %>% 
  mutate(across(-c(Year, Min), ~ round(.x, 2)))

summary_table[1,1] <- '2000'
summary_table[2,1] <- '2024'

# Identify numeric rows for 2000 and 2024
row_2000 <- summary_table %>% filter(Year == "2000") %>% select(-Year)
row_2024 <- summary_table %>% filter(Year == "2024") %>% select(-Year)

# Compute differences
difference_values <- row_2000 - row_2024

# Build difference row
difference_row <- difference_values %>%
  mutate(Year = "Difference") %>%
  relocate(Year)

# Append to table
summary_table <- bind_rows(summary_table, difference_row)


summary_table <- summary_table %>% gt() %>% fmt_number(columns = 'Min', decimals = 5) %>%
  cols_label(Year ~ '') %>%
  tab_header(title = 'Summary Statistics (m)') %>% 
  tab_style(style = cell_text(weight = 'bold'), locations = cells_body(columns = Year)) %>%
  opt_stylize(style =1, color = 'gray')

gtsave(summary_table, r'{E:\01_PROJECTS\01_Roads\05_viz\summary_stats1.png}')

```

```{r fig.height =8, fig.width=10}
#| fig-cap: 'Distribution of Paired Sample Points'

# Sample jitter data separately
jitter_sample <- long_data %>%
  filter(measure_type %in% c('dist_00', 'dist_24')) %>%
  mutate(measure_type = factor(measure_type, levels = c('dist_00', 'dist_24'))) %>%
  filter(
    meters >= quantile(meters, 0.05, na.rm = TRUE),
    meters <= quantile(meters, 0.95, na.rm = TRUE)
  ) %>%
  sample_n(2000)  # Smaller sample just for jitter
levels(jitter_sample$measure_type) <- c("2000", "2024")

# Full data for boxplot
box_data <- long_data %>%
  filter(measure_type %in% c('dist_00', 'dist_24')) %>%
  mutate(measure_type = factor(measure_type, levels = c('dist_00', 'dist_24'))) %>%
  filter(
    meters >= quantile(meters, 0.05, na.rm = TRUE),
    meters <= quantile(meters, 0.95, na.rm = TRUE)
  )
levels(box_data$measure_type) <- c("2000", "2024")

box_plot <- ggplot(box_data, aes(x = measure_type, y = meters)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    data = jitter_sample, 
    aes(x = measure_type, y = meters, color = measure_type), # map color to measure_type
    width = 0.35, alpha = 0.4, size = 4
  ) +
  scale_color_manual(values = c("2000" = "#ea5f94", "2024" = "#3940bb")) + # custom colors
  theme_minimal() +
  xlab("Year") +
  ylab("Distance from Nearest Road (meters)") +
  theme(legend.position = "none")  # optional: remove legend
```

```{r fig.height =8, fig.width=10}
#| fig-cap: "Distribution of Paired Points"
pairwise_histogram <- long_data %>%
  filter(measure_type %in% c('dist_00','dist_24'
  )) %>%
  mutate(measure_type = factor(measure_type,
                          levels = c("dist_00", "dist_24"),
                          labels = c("2000", "2024"))) %>%
  ggplot(mapping = aes(x = log10(meters), fill = measure_type)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 500) +
  scale_fill_manual(values = c("2000" = "#ea5f94", "2024" = "#3940bb")) +
    labs(
    fill = "Year",  # <–– legend title
    x = "Distance to Nearest Road (log10 meters)",
    y = "Count"
  ) +
  theme_minimal() 

```

```{r}
valleys_coves <- long_data %>%
  filter(measure_type %in% c('dist_00','dist_24'
  )) %>%
  filter(ecoregion == "Limestone Valleys and Coves"
  ) %>%
  mutate(measure_type = factor(measure_type,
                          levels = c("dist_00", "dist_24"),
                          labels = c("2000", "2024"))) %>%
  ggplot(mapping = aes(x = meters, fill = measure_type)) +
  geom_density(position = "identity", alpha = 0.5) +
  scale_fill_manual(name = 'Year', values = c("2000" = "#ea5f94", "2024" = "#3940bb")) +
  theme_minimal() +
  xlab('Distance from Nearest Road (m)') +
  ylab('')

```

```{r fig.height =8, fig.width=10}
#| lightbox: true
library(ggplot2)
library(ggridges)
library(dplyr)
library(forcats)

# Combine and prepare
long_data_viz <- long_data %>%
  filter(measure_type %in% c("dist_00", "dist_24")) %>%
  mutate(
    period = case_when(
      measure_type == "dist_00" ~ "2000",
      measure_type == "dist_24" ~ "2024"
    ),
    lulc = fct_reorder(lulc, desc(meters), .fun = mean, na.rm = TRUE)
  )

# Plot
lulc_2000_vs_2024 <- ggplot(long_data_viz, aes(x = log1p(meters), y = lulc, fill = period)) +
  geom_density_ridges(
    alpha = 0.6,
    scale = 0.97,
    stat = "binline",
    bins = 100,
    position = "identity",
    color = NA
  ) +
  theme_ridges() +
  ylab("Land Use/Land Cover") +
  xlab("log1p(Distance in meters)") +
  labs(fill = "Year") + 
scale_fill_manual(name = 'Year', values = c("2000" = "#ea5f94", "2024" = "#3940bb"
)) + 
  ggtitle("Distribution of Distances (2000 vs 2024)")

```

```{r}
#| fig-cap: 'Uniform Random Sampling of 1e6 Points Across Virginia (5,000 shown for interpretability)'
#| echo: false
state_outline <- st_read(r"{E:\01_PROJECTS\01_Roads\02_vector_data\tl_2024_us_state.shp}") %>% filter(GEOID == 51) %>% st_geometry()
sample_pts <- sample_n(all_data, 5000) %>% st_geometry() %>% st_transform(crs = crs(state_outline))

study_area <- ggplot() +
  geom_sf(data = state_outline, fill = NA, color = "black") +
  geom_sf(data = sample_pts, color = "black", size = 2) +
  coord_sf(expand = FALSE, default_crs = sf::st_crs(state_outline)) +
  theme(aspect.ratio = NULL) +
  theme_map()


study_area
```

```{r}

pairwise_histogram / (box_plot + valleys_coves + lulc_2000_vs_2024)

```

## Abbreviations

-   **3DHP** 3D Hydrography Program
-   **EPA** Environmental Protection Agency
-   **GDAL** Geospatial Data Abstraction Library
-   **HUC** Hydrologic Unit Code
-   **NLCD** National Land Cover Database
-   **TIGER** Topologically Integrated Geographic Encoding and Referencing
-   **USGS** United States Geoglogical Survey

# References